%=========================================================================
% (c) Michal Bidlo, Bohuslav Křena, 2008

\include{Preface}

\include{evolutionary-algorithms}

\include
\chapter{Evolutionary algorithms}
\begin{quotation}
	\textit{‘Owing to this struggle for life, variations, however slight and from whatever cause proceeding, if they be in any degree profitable to the individuals of a species, in their infinitely complex relations to other organic beings and to their physical conditions of life, will tend to the preservation of such individuals, and will generally be inherited by the offspring. The offspring, also, will thus have a better chance of surviving, for, of the many individuals of any species which are periodically born, but a small number can survive. I have called this principle, by which each slight variation, if useful, is preserved, by the term Natural Selection’.} (Darwin, 1859 [127], p. 115)
\end{quotation}

In biological evolution, species are selected depending on their relative success in surviving and reproducing in the environment and mutations play a crucial role in this process as they are the key to the adaptation of the species. This concept has been used as a metaphor in computer science and it inspired the formation of a whole new area of artificial intelligence called Evolutionary Computation. This area often deals with an \textit{optimization problem} which can be expressed in the following way: We have a function $f: A \to \mathbb{R}$ from a set $A$ to the real numbers and we are looking for an element $x_0$ in $A$ such that $f(x_0) \leq f(x)$ or $f(x_0) \geq f(x)$ for all $x$ in $A$ (we are looking for either the global minimum or maximum of the function $f$). The domain $A$ of $f$ is called the \textit{search space} and the elements of $A$ are called \textit{candidate solutions}. The function $f$ can be called variously (an \textit{objective function}, a \textit{loss function}), in this text we will call the function $f$ a \textit{fitness function}. A candidate solution that gives us the global maximum or minimum of the fitness function is called an \textit{optimal solution}.
A simple approach to find the optimal solution would be to scour the whole search space and try every candidate solution. This approach will always give us the optimal solution, but the search space can be so vast, that we are not able to try every candidate solution in reasonable time. 
A more sophisticated approach would be to use the technique presented in algorithm \ref{evolutionary-algorithm}. The \textit{population} would be a subset of $A$ (the search space) and every \textit{individual} of the population would represent one element of $A$. The individuals can be also called \textit{chromosomes} in certain types of evolutionary algorithms. This approach helps us to avoid thouse those elements in the search space which wouldn't provide any good solution, but the finding of the optimal solution is not guaranteed.
From the biological point of view, this algorithm can be seen as the driving force behind evolution and from the mathematical point of view it can be seen as a stochastic numerical method for finding global optima of functions that are too hard or impossible to find analytically.
\begin{algorithm}
	\caption{Evolutionary algorithm}\label{evolutionary-algorithm}
	\begin{algorithmic}[1]
		\State Create a population of randomly selected candidate solutions;
		\While{<optimal solution $\notin$ population>}
		\State (Selection) Select the appropriate individuals (parents) for breeding from the current population;
		\State (Mutation) Generate new individuals (offspring) from these parents;
		\State (Reproduction) Replace some or all of the current population with the new individuals;
		\EndWhile
	\end{algorithmic}
\end{algorithm}

\section{Evolutionary Strategies}


\section{Fitness function}
A fitness function provides us means by which we can determine how good the evolved solution is, so that we can distinguish the more appropriate candidates from the less appropriate ones. It is important to mention that the quality of this function is essential for every evolutionary algorithm, because the ability of the algorithm to find the most suitable solution is highly dependent on the ability of the fitness function to distinguish between the individuals. The value of the function is a positive real number in our case and the exact value is not very important, because the algorithm uses this value only for comparing the individuals between each other and we consider those individuals with the lower value as the better ones. There are three fitness functions implemented in this thesis.
\subsection{Best match with the reference solution}
This fitness function was used as the first solution in order to find out if the evolution has the ability to find the desired solution. It is based on the method of least squares which is used in regression analysis. 
In terms of fitness function, every individual is represented by a vector 

%=========================================================================
